---
title: "Algorithms in the Nashlib set in various programming languages -- Part 3"
author: 
   - John C Nash, retired professor, University of Ottawa
   - Peter Olsen, retired ??
date: "11/01/2021"
output: 
  pdf_document:
    keep_tex: false
    toc: true
bibliography: nashlib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## ??require(bookdown) # language engine to display text??
```

# Abstract

Algorithms 16-23 from the book @jncnm79 are implemented in a variety of programming
languages including Fortran, BASIC, Pascal, Python and R. 
These concern rootfinding, function minimisation and
nonlinear least squares.

\pagebreak
# Overview of this document

This section is repeated for each of the parts of
Nashlib documentation.

A companion document **Overview of Nashlib and its Implementations** describes the
process and computing environments for the implementation of Nashlib algorithms.
This document gives comments and/or details relating to implementations of
the algorithms themselves.

Note that some discussion of the reasoning behind certain choices in algorithms or
implementations are given in the Overview document.

\pagebreak
# Algorithm 16 -- Grid search

Grid search -- establishing a regular pattern of parameter values for one or
more arguments of a function and then evaluating that function on the "grid" --
is a brute force approach to finding roots, minima, maxima and other features of a 
function surface. While it cannot be recommended as an efficient method for
finding roots or minima, it offers a way to generate data for plotting the function
surface and for localizing roots or minima when these are not unique. Furthermore,
it is readily understood, and offers a useful starting point in presenting and
understanding a problem.

## Fortran

### Listing

```{r code=xfun::read_utf8('../fortran/alg16.f'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash run16}
gfortran ../fortran/dr16.f
mv ./a.out ../fortran/dr16f.run
../fortran/dr16f.run < ../fortran/dr16f.in
```

<!-- ## BASIC -->

## Pascal

### Listing

```{r code=xfun::read_utf8('../Pascal2021/alg16.pas'), echo=TRUE, eval=FALSE}
```


### Example output

The driver for presenting the example of the Pascal version of Algorithm 16 is
combined with that of Algorithm 17 below.

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

# Algorithm 17 -- Minimize a function of one parameter

It is helpful to be able to visualize a one-parameter function before
trying to find a minimum. R provides a nice way to do this, and also
provides (via the `Brent` method of `optim()`) a way to seek a local
minimum, though we need to provide lower and upper bounds. The original
Algorithm 17 from Nashlib uses a starting guess and a starting stepsize, which
leads to a different approach to finding a minimum. However, the upper and
lower bound approach was used in the 1990 Second Edition and its Turbo Pascal
variant of the code. 

```{r a17ex, echo=TRUE}
 cubfn <- function(x) { x*(x*x-2)-5}
 curve(cubfn, from=-2, to=2)
 res <- optim(par=0.0, fn=cubfn, method="Brent", lower=c(0), upper=c(1))
 cat("Minimum proposed is f(",res$par,")=",res$value,"\n")
```

## Fortran

### Listing

```{r code=xfun::read_utf8('../fortran/alg17.f'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash run17f}
gfortran ../fortran/dr17.f
mv ./a.out ../fortran/dr17f.run
../fortran/dr17f.run < ../fortran/dr17f.in
```

<!-- ## BASIC -->

## Pascal

### Listing

```{r code=xfun::read_utf8('../Pascal2021/alg17.pas'), echo=TRUE, eval=FALSE}
```

### Example output

First we compile the codes.

```{bash script1617cpas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr1617.pas
# copy to run file
mv ../Pascal2021/dr1617 ../Pascal2021/dr1617p.run
```

Then we run the grid search (Algorithm 16) followed by the line search routine
(Algorithm 18).

```{bash script1617prun, eval=TRUE, echo=TRUE}
../Pascal2021/dr1617p.run <../Pascal2021/dr1617p.in >../Pascal2021/dr1617p.out
```


```{r code=xfun::read_utf8('../Pascal2021/dr1617p.out'), echo=TRUE, eval=FALSE}
```

But we can run just the minimizer. Note that above we use only 13 function evaluations
in the minimizer, but now use 17 (for the input used in this example). However, the
grid search used 11 function evaluations prior to the call to the minimizer for a 
total of 24.

```{bash script17prun, eval=TRUE, echo=TRUE}
../Pascal2021/dr1617p.run <../Pascal2021/dr17p.in >../Pascal2021/dr17p.out
```


```{r code=xfun::read_utf8('../Pascal2021/dr17p.out'), echo=TRUE, eval=FALSE}
```


<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

# Algorithm 18 -- Roots of a function of one parameter

We use the same cubic polynomial for our rootfinding test as for the 1D minimizer
(Algorithm 17). R has a built-in 1D rootfinder, `uniroot`. This uses ideas in 
@brent73. As of UseR!2011 in Warwick, the R multiple-precision
package `Rmpfr` (@Rmpfr) did not have a rootfinder because it needed to have
a pure-R code to extend the precision. During a quite period of the conference, 
the author (JN) translated the C code of `uniroot` to plain R, and it is now
the `unirootR` function of `Rmpfr`. The code is in the `rootoned` package at
http://download.r-forge.r-project.org/src/contrib/rootoned_2018-8.28.tar.gz.

Note that this is a different algorithm to that in Nashlib. Moreover, even the
Nashlib codes are not necessarily fully equivalent, as over time minor variations
have crept in. We are also fairly certain that the ideas of Algorithm 18 are NOT
the best for performance. They were written initially for the Data General NOVA
which had very poor quality floating point (24 bit mantissa, likely no guard
digit, no double precision), and with very limited storage. Thus the programming 
goal was reliability rather than efficiency.

```{r a18ex, echo=TRUE}
 cubfn <- function(x) { x*(x*x-2)-5}
## curve(cubfn, from=-2, to=2)
cat("The first attempt fails -- see the plot of the function above.\n")
 res <- try(uniroot(f=cubfn, lower=0, upper=1))
 res <- try(uniroot(f=cubfn, lower=-3, upper=3))
 cat("Root proposed is f(",res$root,")=",res$f.root,"\n")
 cat("Tighter tolerance?\n")
 res <- try(uniroot(f=cubfn, lower=-3, upper=3, tol=1e-10))
 cat("Root proposed is f(",res$root,")=",res$f.root,"\n")
```
## Fortran

### Listing

```{r code=xfun::read_utf8('../fortran/alg17.f'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash run18f}
gfortran ../fortran/dr18.f
mv ./a.out ../fortran/dr18f.run
../fortran/dr18f.run < ../fortran/dr18f.in
```

## BASIC

The code used here was edited from one dated August 30, 1976. Changes were needed
to adapt to the changed syntax of the `PRINT` statement and to allow us to run 
the program inside a scripted environment, but the logic is unchanged. For example,
we have artificially inserted a working set of values to start the Bisection / False
Position rootfinder after the grid search. The original program was designed to 
present the grid search so that the user could interactively choose an interval
for which the endpoints had different function values to start the rootfinder.

### Listing

```{r code=xfun::read_utf8('../BASIC/a18roo.bas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash run18b}
bwbasic ../BASIC/a18roo.bas
```

## Pascal

### Listing

Note that in this routine, we use bisection every 5 function evaluations. That is,
we fix the `nbis` variable at 5. This could easily be changed to make it an input
quantity. 

?? Do we want to discuss why this may be useful?

```{r code=xfun::read_utf8('../Pascal2021/alg18.pas'), echo=TRUE, eval=FALSE}
```

### Example output

First we compile the codes.

```{bash script18pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr1618.pas
# copy to run file
mv ../Pascal2021/dr1618 ../Pascal2021/dr1618p.run
```

Then we run the grid search (Algorithm 16) followed by the Bisection / False position routine
(Algorithm 18).

```{bash script1618aprun, eval=TRUE, echo=TRUE}
../Pascal2021/dr1618p.run <../Pascal2021/dr1618a.in >../Pascal2021/dr1618a.out
```


```{r code=xfun::read_utf8('../Pascal2021/dr1618a.out'), echo=TRUE, eval=FALSE}
```

Let us try WITHOUT grid search first. 

```{bash script1616bprun, eval=TRUE, echo=TRUE}
../Pascal2021/dr1618p.run <../Pascal2021/dr1618b.in >../Pascal2021/dr1618b.out
```

But we can run just the minimizer. Note that above we use only 13 function evaluations
in the minimizer, but now use 17 (for the input used in this example). However, the
grid search used 11 function evaluations prior to the call to the minimizer for a 
total of 24.

```{r code=xfun::read_utf8('../Pascal2021/dr1618b.out'), echo=TRUE, eval=FALSE}
```

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->


# Algorithm 19 and 20 -- Nelder-Mead minimization and Axial search

## Fortran

Algorithm 19 is the Nelder-Mead Simplex direct search minimization. It is perhaps
better called the Polytope method to avoid confusion with the Dantzig Simplex 
method for linear programming. Algorithm 20 is an axial search to escape local 
minima, or possibly the possible collapse of the polytope and a false convergence.
The axial search has no guarantee of success, but it does sometimes avoid false
results.

### Listing - Algorithm 19

```{r code=xfun::read_utf8('../fortran/alg19.f'), echo=TRUE, eval=FALSE}
```

### Listing - Algorithm 20

```{r code=xfun::read_utf8('../fortran/alg20.f'), echo=TRUE, eval=FALSE}
```

### Example output

The example is the Wood4 test function.

```{bash run1920f, eval=TRUE, echo=TRUE}
gfortran ../fortran/dr1920f.f
# copy to run file
mv ./a.out ../fortran/dr1920f.run
../fortran/dr1920f.run <../fortran/dr1920f.in
```


<!-- ## BASIC -->

<!-- ## Pascal -->

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

# Algorithm 21 -- Variable metric minimization

## Fortran

### Listing

```{r code=xfun::read_utf8('../fortran/alg21.f'), echo=TRUE, eval=FALSE}
```

### Example output

We use the WOOD4 function from @jnmws87, page 421 from different starting points. 
The code is set up to return with a negative count of function evaluations
if the pre-set limit is reached. This is tested in the first example case.

```{bash run21f}
gfortran ../fortran/dr21f.f
mv ./a.out ../fortran/dr21f.run
../fortran/dr21f.run < ../fortran/dr21f.in
```

<!-- ## BASIC -->

## Pascal

### Listing

```{r code=xfun::read_utf8('../Pascal2021/alg21.pas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash script21pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr21p.pas
# copy to run file
mv ../Pascal2021/dr21p ../Pascal2021/dr21p.run
../Pascal2021/dr21p.run >../Pascal2021/dr21p.out
```


```{r code=xfun::read_utf8('../Pascal2021/dr21p.out'), echo=TRUE, eval=FALSE}
```


<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

# Algorithm 22 -- Conjugate gradients minimizers

## Fortran

### Listing

```{r code=xfun::read_utf8('../fortran/alg22.f'), echo=TRUE, eval=FALSE}
```

### Example output

We use the WOOD4 function from @jnmws87, page 421 from different starting points. 
The code is set up to return with a negative count of function evaluations
if the pre-set limit is reached.

??? need to explain and maybe clean up output

```{bash run22f}
gfortran ../fortran/dr22f.f
mv ./a.out ../fortran/dr22f.run
../fortran/dr22f.run < ../fortran/dr22f.in
```


## BASIC

The original BASIC version of Algorithm 22 seems to have been lost to time. However,
the work towards @jnmws87 created a new Conjugate Gradients minimizer based on Algorithm
22 but with the addition of code allowing both bounds and masks constraints. The code
is available from archive.org at ??


## Pascal

### Listing

```{r code=xfun::read_utf8('../Pascal2021/alg22.pas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash script22pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr22p.pas
# copy to run file
mv ../Pascal2021/dr22p ../Pascal2021/dr22p.run
../Pascal2021/dr22p.run <../Pascal2021/dr22p.in >../Pascal2021/dr22p.out
```


```{r code=xfun::read_utf8('../Pascal2021/dr22p.out'), echo=TRUE, eval=FALSE}
```




<!-- ## Python -->

## R

In the early 2000s, one of us (JN) tried implementing his optimization algorithms
from Nashlib in R. The outcome of this exercise, coloured by collaborations with 
other workers (in particular Ravi Varadhan, Hans Wener Borchers, Ben Bolker, Duncan
Murdoch, and Gabor Grothendiek), is the R-project package `optimx`. However, in 
considering Algorithm 22, which which I have never been happy, I came across 
@Dai99, where a very small change in the logic of Algorithm 22 allowed the three
search direction updates offered separately in Algorithm 22 to be very elegantly
combined. The resulting program, first packaged as R package `Rcgmin` before it
was subsumed into `optimx` in 2019, works much, much better than the original
Algorithm 22. 


<!-- ## Others -->

# Algorithm 23 -- Marquardt method for nonlinear least squares

## Fortran

### Listing

```{r code=xfun::read_utf8('../fortran/alg23.f'), echo=TRUE, eval=FALSE}
```

### Example output

We again test the code with a rather nasty 4-parameter problem called WOOD4
(@jnmws87, page 421) from different starting points. An example is 
included with a function evaluation limit that forces early termination.
Caution!!


```{bash run23f}
gfortran ../fortran/a23.f
mv ./a.out ../fortran/a23f.run
../fortran/a23f.run < ../fortran/dr23f.in
```


## BASIC

### Listing


<!-- ```{r code=xfun::read_utf8('../BASIC/'), echo=TRUE, eval=FALSE} -->
<!-- ``` -->

### Example output

<!-- ```{bash script23pas, eval=TRUE, echo=TRUE} -->
<!-- fpc ../Pascal2021/dr23p.pas -->
<!-- # copy to run file -->
<!-- mv ../Pascal2021/dr23p ../Pascal2021/dr23p.run -->
<!-- ../Pascal2021/dr23p.run  >../Pascal2021/dr23p.out -->
<!-- ``` -->


## Pascal

### Listing


```{r code=xfun::read_utf8('../Pascal2021/alg23.pas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash script23pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr23p.pas
# copy to run file
mv ../Pascal2021/dr23p ../Pascal2021/dr23p.run
../Pascal2021/dr23p.run  >../Pascal2021/dr23p.out
```

```{r code=xfun::read_utf8('../Pascal2021/dr23p.out'), echo=TRUE, eval=FALSE}
```

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

\pagebreak
# Algorithm 27 -- Hooke and Jeeves pattern search minimization

<!-- ## Fortran -->

## BASIC

The Hooke and Jeeves code below is a modified version of that in @nash90NLEX. 

### Listing

```{r code=xfun::read_utf8('../BASIC/hj27.bas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash run27b}
bas ../BASIC/hj27.bas <../BASIC/yes.in
```


## Pascal

### Listing

```{r code=xfun::read_utf8('../Pascal2021/alg27.pas'), echo=TRUE, eval=FALSE}
```

### Example output

### Example output

Use Rosenbrock banana-shaped valley problem in 2 dimensions.

```{bash script27pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr27.pas
# copy to run file
mv ../Pascal2021/dr27 ../Pascal2021/dr27.run
../Pascal2021/dr27.run  >../Pascal2021/dr27p.out
```

```{r code=xfun::read_utf8('../Pascal2021/dr27p.out'), echo=TRUE, eval=FALSE}
```

<!-- ## Python -->

## R

### Listing

```{r code=xfun::read_utf8('../R/hjn.R'), echo=TRUE, eval=FALSE}
```

### Example output

We use the Rosenbrock banana-shaped valley problem in 2 dimensions.

```{r script27R, eval=TRUE, echo=TRUE}
source("../R/hjn.R") # bring the Hooke and Jeeves minimizer code into workspace
fminfn <-function(x){ 
    val<-((x[2]-x[1]^2)^2)*100.0+(1.0-x[1])^2
    val
}
x0<-c(-1.2,1)
cat("Check fn at c(-1.2,1)=",fminfn(x0),"\n")
rslt <- hjn(par=x0, fn=fminfn, control=list(trace=0))
print(rslt)
```


<!-- ## Others -->




<!-- # Cleanup of working files -->

<!-- The following script is included to remove files created during compilation or -->
<!-- execution of the examples. -->

```{bash cleanup, echo=FALSE, eval=TRUE}
## remove object and run files
cd ../fortran/
echo `pwd`
rm  *.o
rm  *.run
rm *.out
cd ../Pascal2021/
echo `pwd`
rm  *.o
rm  *.run
rm *.out
cd ../BASIC
echo `pwd`
rm *.out
cd ../Documentation
## ?? others
```


# References