---
title: "Algorithms in the Nashlib set in various programming languages -- Part 4"
author: 
   - John C Nash, retired professor, University of Ottawa
   - Peter Olsen, retired ??
date: "11/01/2021"
output: 
  pdf_document:
    keep_tex: false
    toc: true
bibliography: nashlib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## ??require(bookdown) # language engine to display text??
```

# Abstract

Algorithms 24 and 25 from the book @jncnm79 are implemented in a variety of programming
languages including Fortran, BASIC, Pascal, Python and R. These routines concern the 
use of iterative methods, in particular conjugate gradients, to solve linear algebra
problems. 

\pagebreak
# Overview of this document

This section is repeated for each of the parts of
Nashlib documentation.

A companion document **Overview of Nashlib and its Implementations** describes the
process and computing environments for the implementation of Nashlib algorithms.
This document gives comments and/or details relating to implementations of
the algorithms themselves.

Note that some discussion of the reasoning behind certain choices in algorithms or
implementations are given in the Overview document.

\pagebreak



# Algorithm 24 -- conjugate gradients for linear equations and least squares

## Fortran

### Listing

```{r code=xfun::read_utf8('../fortran/alg24.f'), echo=TRUE, eval=FALSE}
```

### Example output

We solve the linear equations problem 

$$ A x = b$$
for $A$ being the Frank matrix and $b$ the vector that is all $A 1$, that is, $A$
times a vector of ones. Different orders
of problem are tried. This problem has an obvious solution, which is that

$$ x_i = 1 $$
and the elements of $A$ are integers, so we can compute the error in the solution
as well as the residual 

$$ b - A x $$


```{r chkfrankle}
A <- matrix(0, nrow=5, ncol=5)
b <- rep(1,5)

for (i in 1:5) {
   for (j in 1:5){
       A[i,j]=min(i,j)
   }
}
print(A)
C <- A %*% b
print(C)
```



Note that the program, as given, is single precision, so some
of the results are not terribly accurate. Also the control `IMULT` is an upper 
limit on the number of matrix multiplications before the process is artificially
halted. We still get an approximate solution at this stage, and the result
accuracy in terms of residuals and error.

```{bash run24le}
gfortran ../fortran/dr24.f
mv ./a.out ../fortran/dr24le.run
../fortran/dr24le.run < ../fortran/a24f.in
```

<!-- ## BASIC -->

## Pascal

### Listing of Algorithm 24


### Example output for linear equations

Algorithm 24 solves linear equations, so no extra code is presented.

```{bash pas24le}
fpc ../Pascal2021/dr24le.pas
mv ../Pascal2021/dr24le ../Pascal2021/dr24le.run
../Pascal2021/dr24le.run <../Pascal2021/dr24le.in
```

### Example output for least squares

```{bash pas24ls}
fpc ../Pascal2021/dr24ls.pas
mv ../Pascal2021/dr24ls ../Pascal2021/dr24ls.run
../Pascal2021/dr24ls.run <../Pascal2021/dr24ls.in
```

### Example output for inverse iteration for eigensolutions

We need extra code ?? Algorithm 24 solves linear equations, so no extra code is presented.






<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

\pagebreak
# Algorithm 25 -- Rayleight quotient minimization

## Fortran 

### Listing

```{r code=xfun::read_utf8('../fortran/a25.f'), echo=TRUE, eval=FALSE}
```

### Example output

?? explanation needed

```{bash run25}
gfortran ../fortran/a25.f
mv ./a.out ../fortran/a25.run
../fortran/a25.run < ../fortran/a25.in
```


<!-- ## BASIC -->

## Pascal

### Listing

```{r code=xfun::read_utf8('../Pascal2021/alg25.pas'), echo=TRUE, eval=FALSE}
```

### Example output

We use the same example as for Algorithm 15, with $A$ the unit matrix and $B$ the
Frank matrix, here of order 5.

Note that we could modify the program to work with $-A$ to get the negative of the
largest eigenvalue. Various shifting strategies might be used to get other solutions,
but they are rather inconvenient and not recommended.

```{bash script25pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr25.pas
# copy to run file
mv ../Pascal2021/dr25 ../Pascal2021/dr25.run
../Pascal2021/dr25.run <../Pascal2021/dr25p.in >../Pascal2021/dr25p.out
```

```{r code=xfun::read_utf8('../Pascal2021/dr25p.out'), echo=TRUE, eval=FALSE}
```


<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->



# Cleanup of working files

<!-- The following script is included to remove files created during compilation or -->
<!-- execution of the examples. -->

```{bash cleanup, echo=FALSE}
## remove object and run files
cd ../fortran/
echo `pwd`
rm  *.o
rm  *.run
rm *.out
cd ../Pascal2021/
echo `pwd`
rm  *.o
rm  *.run
rm *.out
cd ../BASIC
echo `pwd`
rm *.out
cd ../Documentation
## ?? others
```


# References