---
title: "Algorithms in the Nashlib set in various programming languages -- Part 1"
author: 
   - John C Nash, retired professor, University of Ottawa
   - Peter Olsen, retired ??
date: "11/01/2021"
output: 
  pdf_document:
    keep_tex: false
    toc: true
bibliography: nashlib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## ??require(bookdown) # language engine to display text??
```

# Abstract

Algorithms from the book @jncnm79 are implemented in a variety of programming
languages including Fortran, BASIC, Pascal, Python and R. 

\pagebreak
# Overview of this document

A companion document **Overview of Nashlib and its Implementations** describes the
process and computing environments for the implementation of Nashlib algorithms.
This document gives comments and/or details relating to implementations of
the algorithms themselves.

Note that some discussion of the reasoning behind certain choices in algorithms or
implementations are given in the Overview document.

\pagebreak
# Algorithms 1 and 2 -- one-sided SVD and least squares solution

These were two of the first algorithms to interest the first author in compact codes.
At the time (1973-1978) he was working at Agriculture Canada in support of econometric
modeling. More or less "regular" computers required accounts linked to official projects,
but there was a time-shared Data General NOVA that offered 4K to 7K byte working spaces
for data and programs in interpreted BASIC. BASIC of a very similar dialect was available
also on an HP 9830 calculator. On these machines, availability of a terminal or the
calculator was the only limitation to experimentation with recent innovations in
algorithms. In particular, a lot of modeling was done with linear least squares
regression, mostly using the traditional normal equations. The singular value decomposition
and other methods such as the Householder, Givens or Gram-Schmidt 
approaches to the QR matrix decomposition were relatively recent innovations. However,
the code for the Golub-Kahan SVD was rather long for both the hardware and the 
BASIC language. Instead, a one-sided Jacobi method was developed from ideas of @Hestenes58
and @Chartres62. Some work by @Kaiser72 was also observed. Later workers have generally credited 
Hestenes with this approach, and
he certainly wrote about it, but we (JN) suspect strongly that he never actually
attempted an implementation. In a conversation at a conference, Chartres said that some 
experiments were tried, but that he believed no production usage occurred. We must
remember that access to computers until the 1970s was quite difficult. 

The method published in @jn75cj and later revised in @jnss87 ignored some advice that Jacobi
rotations should not use angles greater than $\pi/4$ (see @ForsytheHenrici1960). This allowed of a cyclic process that
not only developed a form of the decomposition, but also sorted it to effectively present
the singular values in descending order of size. This avoided extra program code of about
half the length of the svd routine.

About 2 decades after @jn75cj, there was renewed interest in one-sided Jacobi methods, but 
rather little acknowledgment of the earlier development, and much more complicated codes.
?? How far to reference more recent developments??


## Fortran

Note that these are single precision codes. Very few modern calculations are carried out at
this precision. Moreover, the dialect of Fortran (Fortran 77) is now decidedly old-fashioned, 
though it compiles and executes just fine.


### Listing -- Algorithm 1

```{r code=xfun::read_utf8('../fortran/alg01.f'), echo=TRUE, eval=FALSE}
```

### Listing -- Algorithm 2

```{r code=xfun::read_utf8('../fortran/alg02.f'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash run0102}
gfortran ../fortran/dr0102.f
mv ./a.out ../fortran/dr0102.run
../fortran/dr0102.run < ../fortran/dr0102.in
```


### Special implementations

Most singular value decomposition codes are much, much more complicated than
Algorithm 1 of the Nashlib collection. For some work on the magnetic field of
Jupiter for NASA, Sidey Timmins has used an extended (quad) precision version
of the method. One of us (JN) has converted an updated algorithm (@jnss87) to
the Fortran 95 dialect so the multiple precision FM Fortran tools of David M. Smith
(see \url{http://dmsmith.lmu.build/}).

?? include this code and example in the repo??


## BASIC

### Listing

```{r code=xfun::read_utf8('../BASIC/dr0102.bas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash scriptbas}
bwbasic ../BASIC/dr0102.bas
echo "done"
```


## Pascal

### Listing -- Algorithm 1

```{r code=xfun::read_utf8('../Pascal2021/alg01.pas'), echo=TRUE, eval=FALSE}
```

### Listing -- Algorithm 2

```{r code=xfun::read_utf8('../Pascal2021/alg02.pas'), echo=TRUE, eval=FALSE}
```

### Example output

For some reason not yet understood, running the compiled Pascal program does not transfer the output
to our Rmarkdown output, so we resort to saving the output and then listing it as we do program code.

```{bash scriptpas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr0102.pas
mv ../Pascal2021/dr0102 ../Pascal2021/dr0102.run
# now execute it
../Pascal2021/dr0102.run > ../Pascal2021/dr0102.out
```

```{r code=xfun::read_utf8('../Pascal2021/dr0102.out'), echo=TRUE, eval=FALSE}
```
For some reason, we get extra line-feed characters in the output file. They are easily removed
with a text editor from the output file, but their origin is unclear. JN 2021-1-20 ??

## Python

Pending ...

## R

### Listing

While based on @jnss87, the following code shows that R can be used quite easily
to implement Algorithm 1. The least squares solution (Algorithm 2) is embedded
in the example output.

```{r code=xfun::read_utf8('../R/Nashsvd.R'), echo=TRUE, eval=FALSE}
```


### Example output

```{r exa1R, eval=TRUE, echo=TRUE}
# test taken from dr0102.pas
A<-matrix(0, 4,3)
A[1,]<-c(5, 1e-6, 1)
A[2,]<-c(6, 0.999999, 1)
A[3,]<-c(7, 2.00001, 1)
A[4,]<-c(8, 2.9999, 1)
print(A)
b<-c(1,2,3,4)
print(b)
# try the R-base svd 
sA <- svd(A)
sA
yy <- t(sA$u) %*% as.matrix(b)
xx <- sA$v %*% diag(1/sA$d) %*% yy
xx
# Now the Nashsvd code (this is likely NOT true to 1979 code)
source("../R/Nashsvd.R")
nsvd <- Nashsvd(A)
print(nsvd)
# Note least squares solution can be done by matrix multiplication
U <- nsvd$u
V <- nsvd$v
d <- nsvd$d
di <- 1/d
di <- diag(di) # convert to full matrix -- note entry sizes
print(di)
lsol <- t(U) %*% b
lsol <- di %*% lsol
lsol <- V %*% lsol
print(lsol)
res <- b - A %*% lsol
print(res)
cat("sumsquares = ", as.numeric(crossprod(res)))
# now set smallest singular value to 0 and in pseudo-inverse
dix <- di
dix[3,3] <- 0
lsolx <- V %*% dix %*% t(U) %*% b
# this gives a very different least squares solution
print(lsolx)
# but the residuals (in this case) are nearly 0 too
resx <- b - A %*% lsolx
cat("sumsquares = ", as.numeric(crossprod(resx)))

```

## Others

Pending ...

?? Could we f2c the Fortran and manually tweak to get a C code?

There is also a C version in 

\url{https://github.com/LuaDist/gsl/blob/master/linalg/svd.c}

===========================================================================

\pagebreak
# Algorithm 3 -- Givens' decomposition

The Givens and Householder decompositions of a rectangular $m$ by $n$ matrix $A$
($m  >= n$) both give an $m$ by $m$ orthogonal matrix $Q$ and an upper-triangular
$n$ by $n$ matrix $R$ whose product $QR$ is a close approximation of $A$. At the time
@jncnm79 was being prepared, the Givens approach seemed to give a more compact program
code, though neither approach is large.

In practice, if one is trying to solve linear equations 

$$    A  x  =  b $$

or linear least squares problems of the form

$$    A  x  ~=  b  $$

then the right hand side (RHS)  $b$ can be appended to the matrix A so that the
resulting working matrix

$$  W  =  [A | b]  $$
is transformed during the formation of the $Q$ matrix into 

$$   W_{trans}  = [ R | Q' b]  $$
This saves us the effort of multiplying $b$ by the transpose of $Q$ before we 
back-solve for x. 

In fact, $m$ does not have to be greater than or equal to $n$. However, underdetermined
systems of equations do raise some issues that we will not address here.


It is therefore unnecessary to store $Q$, which when @jncnm79 was being prepared was a
potentially large matrix. There are alternative designs of the code which could save
information on the plane rotations that make up $Q$. Such codes can then apply the 
rotations to a unit matrix of the right size to reconstruct $Q$ as needed. However,
these details have largely become irrelevant in an age of cheap memory chips.

### Notes on work in 2021 with Sidey Timmons

Starting in late 2019, one of us (JN) exchanged ideas relating to the one-sided
Jacobi approach to the singular value decomposition. The application was to work
to estimate the magnetic field of Jupiter from spacecraft measurements. The compact
Nash codes were attractive as extremely high precision calculations were needed
that required special programming language processors, for example, FM Fortran
(@Smith91afortran, see also \url{http://dmsmith.lmu.build/}). 

As this work progressed, it was noted that work on LAPACK had, in 2000, revealed 
quite large discrepancies between some implementations of the Givens' method for
the QR decomposition, and a working paper was prepared, eventually published
<<<<<<< HEAD
as @Bindel2002Givens. Some aspects of these ideas will 
be included in this collection. (This comment written as of 2021-3-21.) 
=======
as @Bindel2002Givens. Some aspects of these ideas will (??written as of 2021-3-21)
be included in this collection.
>>>>>>> a42e9e6b62b1fdb1ff5ad262fc7f202f091c83dd


## Fortran

### Listing -- Algorithm 3

```{r code=xfun::read_utf8('../fortran/alg03.f'), echo=TRUE, eval=FALSE}
```

### Example output

The following output presents an example using the Frank matrix as a test. 
As a precaution, we use a 1 by 1 matrix as our first test. We have seen situations 
where otherwise reliable programs have failed on such trivial cases.

```{bash run3}
gfortran ../fortran/dr03.f
mv ./a.out ../fortran/dr03.run
../fortran/dr03.run < ../fortran/a3data.in > ../fortran/a3out.txt
```

```{r code=xfun::read_utf8('../fortran/a3out.txt'), echo=TRUE, eval=FALSE}
```


## BASIC

### Listing

The following listing also uses the Frank matrix as a test. The code has been 
adjusted for fixed input to allow it to be run within the `knitr` processor for
`Rmarkdown`.

```{r code=xfun::read_utf8('../BASIC/a3.bas'), echo=TRUE, eval=FALSE}
```


### Example output

As a precaution, we use a 1 by 1 matrix as our first test. We have seen situations 
where otherwise reliable programs have failed on such trivial cases.

```{bash run3bas}
bwbasic ../BASIC/a3.bas
```


## Pascal

### Listing -- Algorithm 3, column-wise approach

```{r code=xfun::read_utf8('../Pascal2021/alg03.pas'), echo=TRUE, eval=FALSE}
```

### Example output -- column-wise approach

```{bash script56pas, eval=TRUE}
fpc ../Pascal2021/dr03.pas
mv ../Pascal2021/dr03 ../Pascal2021/dr03.run
../Pascal2021/dr03.run >../Pascal2021/dr03.out
```


```{r code=xfun::read_utf8('../Pascal2021/dr03.out'), echo=TRUE, eval=FALSE}
```


<!-- ## Python -->

## R

### Listing -- Algorithm 3, column-wise approach

```{r code=xfun::read_utf8('../R/dr03.R'), echo=TRUE, eval=FALSE}
```

### Example output

Below is a first try that uses explicit loops that are known to be
inefficient in R. In this version, we work across the columns in
the outer loop. For simplicity in running the code within `knitr`,
the input of matrix dimensions has been replaced with simple 
assignments.

```{r code=xfun::read_utf8('../R/dr03.R'), echo=TRUE, eval=TRUE}
```

We can simplify the loops for steps 7 and 8 as follows without change
in the results.

```{r alg3Rimploop, echo=TRUE, eval=FALSE}
#    C  STEP 7
         Pv <- A[j,]
         A[j,] <- C*Pv+S*A[k,]
         A[k,] <- -S*Pv+C*A[k,]
# C  STEP 8
         Pv <- Q[,j]
         Q[,j] <- C*Pv+S*Q[,k]
         Q[,k] <- -S*Pv+C*Q[,k]
```


## Issues with calculating the Givens' plane rotations

@Bindel2002Givens point out `"the
LAPACK [Anderson et al. 1999] routines SLARTG, CLARTG, SLARGV and
CLARGV, the Level 1 BLAS routines SROTG and CROTG [Lawson et al. 1979],
as well as Algorithm 5.1.5 in Golub and Van Loan [1996] can get significantly
different answers for mathematically identical inputs."`

Based on the LAPACK routine `slartg.f`, the following version of Algorithm 3 
incorporates the suggestions of @Bindel2002Givens. 

```{r code=xfun::read_utf8('../R/dr03B.R'), echo=TRUE, eval=FALSE}
```

### Examples of effect of Bindel et al. Givens' plane rotations

?? need some illustrative cases that are not too big.


<!-- ## Others -->

# Algorithm 4 -- Row-oriented SVD and least squares solution

<<<<<<< HEAD
The essence of Algorithm 4 is to use a QR decomposition to triangularize a "long
skinny" matrix, that is, $m$ by $n$ with $m >> n$. The resulting $n$ by $n$ matrix
is then further decomposed to a singular value decomposition.

$$                 A  =  Q R $$

$$                 R  =  P S V^T $$
Thus  

$$                 A  =  Q P S V^T $$
so

$$                 U = Q P $$
There are many possibilities for both the QR and SVD parts of this process. In the
early 1970s, one of us (JN) needed to solve a problem with $m=196$ by $n=25$ columns
(variables plus a constant).
=======
The essence of Algorithm 4
>>>>>>> a42e9e6b62b1fdb1ff5ad262fc7f202f091c83dd


## Fortran

### Listing -- SVD and least squares solution

We include the plane-rotation sub-program.

```{r code=xfun::read_utf8('../fortran/alg04.f'), echo=TRUE, eval=FALSE}
```

```{bash run4}
gfortran ../fortran/dr04.f
mv ./a.out ../fortran/dr04.run
../fortran/dr04.run < ../fortran/dr04f.in > ../fortran/dr04f.out
```

```{r code=xfun::read_utf8('../fortran/dr04f.out'), echo=TRUE, eval=FALSE}
```


<<<<<<< HEAD
=======


>>>>>>> a42e9e6b62b1fdb1ff5ad262fc7f202f091c83dd
### Example output

<!-- ## BASIC -->

<!-- ## Pascal -->

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

\pagebreak
# Algorithms 5 and 6 -- Gaussian elimination and back-solution

## Fortran

### Listing -- Algorithm 5 Gaussian elimination

```{r code=xfun::read_utf8('../fortran/alg05.f'), echo=TRUE, eval=FALSE}
```

### Listing -- Algorithm 6 Back-solution of upper triangular equation systems

```{r code=xfun::read_utf8('../fortran/alg06.f'), echo=TRUE, eval=FALSE}
```


### Example output

```{bash run56}
gfortran ../fortran/dr0506.f
mv ./a.out ../fortran/dr0506.run
../fortran/dr0506.run > ../fortran/dr0506out.txt
```

```{r code=xfun::read_utf8('../fortran/dr0506out.txt'), echo=TRUE, eval=FALSE}
```


<!-- ## BASIC -->

## Pascal

### Listing -- Algorithm 5, column-wise approach

```{r code=xfun::read_utf8('../Pascal2021/alg05.pas'), echo=TRUE, eval=FALSE}
```

### Listing -- Algorithm 6, back-solution of upper triangular equations

```{r code=xfun::read_utf8('../Pascal2021/alg06.pas'), echo=TRUE, eval=FALSE}
```

### Example output -- column-wise approach

```{bash script3pas, eval=TRUE}
fpc ../Pascal2021/dr0506.pas
mv ../Pascal2021/dr0506 ../Pascal2021/dr0506.run
../Pascal2021/dr0506.run >../Pascal2021/dr0506.out
```


```{r code=xfun::read_utf8('../Pascal2021/dr0506.out'), echo=TRUE, eval=FALSE}
```

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

# Algorithms 7 and 8 -- Choleski decomposition and back-solution

## Fortran

### Listing -- Algorithm 7 Choleski decomposition

```{r code=xfun::read_utf8('../fortran/alg07.f'), echo=TRUE, eval=FALSE}
```

### Listing -- Algorithm 8 Choleski Back-solution

```{r code=xfun::read_utf8('../fortran/alg08.f'), echo=TRUE, eval=FALSE}
```



<!-- ## BASIC -->


## Pascal

### Listing -- Algorithm 7 Choleski decomposition

```{r code=xfun::read_utf8('../Pascal2021/alg07.pas'), echo=TRUE, eval=FALSE}
```

### Listing -- Algorithm 8 Choleski Back-solution

```{r code=xfun::read_utf8('../Pascal2021/alg08.pas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash script78pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr0708.pas
# copy to run file
mv ../Pascal2021/dr0708 ../Pascal2021/dr0708.run
../Pascal2021/dr0708.run <../Pascal2021/dr0708p.in >../Pascal2021/dr0708p.out
```

```{r code=xfun::read_utf8('../Pascal2021/dr0708p.out'), echo=TRUE, eval=FALSE}
```


<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

\pagebreak
# Algorithm 9 -- Bauer-Reinsch matrix inversion

@wilkinson1971linear, pages 45-49, is a contribution entitled **Inversion of Positive
Definite Matrices by the Gauss-Jordan Method**. It hardly mentions, but appears to
assume, that the matrix to be inverted is symmetric. Two Algol procedures are provided,
one for a matrix stored as a square array, the other for the a matrix where only the
lower triangle is stored as a single vector in row-wise order. That is, if A is
of order n=3 and has values

```
  1  2  4
  2  3  5
  4  5  6
```

Then the corresponding vector of 6 = n*(n+1)/2 values is 

```
   1  2  3  4  5  6
```

By some exceedingly clever coding and matrix manipulation, Bauer and Reinsch developed
tiny codes that invert a positive-definite matrix *in situ* using only one extra vector
of length n. Thus, besides the memory to store a very small code, we need only 
n*(n+3)/2 floating point numbers and a few integers to index arrays.

Truthfully, we rarely need an explicit matrix inverse, and the most common positive-definite
symmetric matrix that arises in scientific computations is the sum of squares and cross-products
(SSCP) in the normal equations used for linear (or also nonlinear) least squares problems. However,
the formation of this SSCP matrix is rarely the best approach to solving least squares problems.
The SVD introduced in Algorithm 1 and the least squares solution in Algorithm 2 lead to better
methods. (??mention A4, Choleski in A7, A8 etc.)

Despite these caveats, the Bauer-Reinsch algorithm is interesting 
as a historical curiosity,
showing what can be done when resources are very limited.

## Fortran

### Listing

```{r code=xfun::read_utf8('../fortran/alg09.f'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash run09}
## #!/bin/bash
gfortran ../fortran/a9.f
mv ./a.out ../fortran/a9.run
../fortran/a9.run
```

## BASIC

### Listing

```{r code=xfun::read_utf8('../BASIC/a9.bas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash a9bas}
bwbasic ../BASIC/a9.bas >../BASIC/a9.out
# echo "done"
```

```{r code=xfun::read_utf8('../BASIC/a9.out'), echo=TRUE, eval=FALSE}
```


## Pascal

### Listing

```{r code=xfun::read_utf8('../Pascal2021/alg09.pas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash script9pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr09.pas
# copy to run file
mv ../Pascal2021/dr09 ../Pascal2021/dr09.run
../Pascal2021/dr09.run >../Pascal2021/dr09p.out
```

```{r code=xfun::read_utf8('../Pascal2021/dr09p.out'), echo=TRUE, eval=FALSE}
```

## Python

WARNING: interim test only!!!???

### Listing

The Algorithm 9 code:

```{r code=xfun::read_utf8('../python/alg09.py'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash trya9py}
python3 ../python/dr09py.py
```

## R

### Listing and Example output


```{r code=xfun::read_utf8('../R/dr09.R'), echo=TRUE, eval=TRUE}
```



## Others



# Cleanup of working files

The following script is included to remove files created during compilation or
execution of the examples.

```{bash cleanup, echo=TRUE}
## remove object and run files
cd ../fortran/
echo `pwd`
rm  *.o
rm  *.run
rm *.out
cd ../Pascal2021/
echo `pwd`
rm  *.o
rm  *.run
rm *.out
cd ../BASIC
echo `pwd`
rm *.out
cd ../Documentation
## ?? others
```


# References