---
title: "Algorithms in the Nashlib set in various programming languages -- Part 2"
author: 
   - John C Nash, retired professor, University of Ottawa
   - Peter Olsen, retired ??
date: "11/01/2021"
output: 
  pdf_document:
    keep_tex: false
    toc: true
bibliography: nashlib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## ??require(bookdown) # language engine to display text??
```

# Abstract

Algorithms 10 and 13-15 from the book @jncnm79 are implemented in a variety of programming
languages including Fortran, BASIC, Pascal, Python and R. 
These concern the eigensolutions of a real, symmetric
matrix and its generalization to include a symmetric
positive-definite metric.

\pagebreak
# Overview of this document

This section is repeated for each of the parts of
Nashlib documentation.

A companion document **Overview of Nashlib and its Implementations** describes the
process and computing environments for the implementation of Nashlib algorithms.
This document gives comments and/or details relating to implementations of
the algorithms themselves.

Note that some discussion of the reasoning behind certain choices in algorithms or
implementations are given in the Overview document.

\pagebreak
# Algorithm 10 -- Inverse iteration via Gaussian elimination

The purpose of this algorithm is to find a single eigensolution of
a matrix $A$ via inverse iteration. That is, we want solutions ($e$,$x$)
of 

$$  A x = e x$$
The programs do not require a symmetric matrix, which leaves open the possibility
that a solution may not exist in the unsymmetric case.

## Fortran

The Algorithm 10 code:

```{r code=xfun::read_utf8('../fortran/alg10.f'), echo=TRUE, eval=FALSE}
```

### Example output

We illustrate by finding a single eigensolution of the Hilbert segments of
order 5 and 10. ?? Do we want to swap in the Frank matrix (the computations
are generally easier)?

```{bash run10}
## #!/bin/bash
gfortran ../fortran/d10.f
mv ./a.out ../fortran/d10.run
../fortran/d10.run
```
## BASIC

### Listing

```{r code=xfun::read_utf8('../BASIC/a10.bas'), echo=TRUE, eval=FALSE}
```

### Example output

In this case we use the Frank matrix for our test.

```{bash a10bas}
bwbasic ../BASIC/a10.bas >../BASIC/a10.out
# echo "done"
```

```{r code=xfun::read_utf8('../BASIC/a10.out'), echo=TRUE, eval=FALSE}
```


## Pascal

### Listing

```{r code=xfun::read_utf8('../Pascal2021/alg10.pas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash script10pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr10.pas
# copy to run file
mv ../Pascal2021/dr10 ../Pascal2021/dr10.run
../Pascal2021/dr10.run <../Pascal2021/dr10p.in >../Pascal2021/dr10p.out
```

```{r code=xfun::read_utf8('../Pascal2021/dr10p.out'), echo=TRUE, eval=FALSE}
```

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

\pagebreak
# Algorithm 13

## Fortran

### Listing - Algorithim 13

```{r code=xfun::read_utf8('../fortran/alg13.f'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash run13}
## #!/bin/bash
gfortran ../fortran/dr13.f
mv ./a.out ../fortran/dr13.run
../fortran/dr13.run < ../fortran/dr13.in
```

## BASIC

### Listing

```{r code=xfun::read_utf8('../BASIC/a13.bas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash a13bas}
bwbasic ../BASIC/a13.bas  >../BASIC/a13.out
# echo "done"
```

```{r code=xfun::read_utf8('../BASIC/a13.out'), echo=TRUE, eval=FALSE}
```

## Pascal

### Listing -- Algorithm 13

```{r code=xfun::read_utf8('../Pascal2021/alg13.pas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash script13pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr13.pas
# copy to run file
mv ../Pascal2021/dr13 ../Pascal2021/dr13.run
../Pascal2021/dr13.run <../Pascal2021/dr13p.in >../Pascal2021/dr13p.out
```

```{r code=xfun::read_utf8('../Pascal2021/dr13p.out'), echo=TRUE, eval=FALSE}
```

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

\pagebreak
# Algorithm 14 -- Jacobi symmetric matrix eigensolutions

## Fortran

### Listing -- Algorithm 14

```{r code=xfun::read_utf8('../fortran/alg14.f'), echo=TRUE, eval=FALSE}
```

### Example output

We use a Moler matrix of order 5 for our example.

```{bash run14f}
## #!/bin/bash
gfortran ../fortran/dr14x.f
mv ./a.out ../fortran/dr14x.run
../fortran/dr14x.run < ../fortran/dr14xf.in
```

<!-- ## BASIC -->

<!-- ### Listing -->

<!-- ```{r code=xfun::read_utf8('../BASIC/a13.bas'), echo=TRUE, eval=FALSE} -->
<!-- ``` -->

<!-- ### Example output -->

<!-- ```{bash a13bas} -->
<!-- bwbasic ../BASIC/a13.bas  >../BASIC/a13.out -->
<!-- # echo "done" -->
<!-- ``` -->

<!-- ```{r code=xfun::read_utf8('../BASIC/a13.out'), echo=TRUE, eval=FALSE} -->
<!-- ``` -->


## Pascal

### Listing -- Algorithm 14

```{r code=xfun::read_utf8('../Pascal2021/alg14.pas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash script14pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr14x.pas
# copy to run file
mv ../Pascal2021/dr14x ../Pascal2021/dr14x.run
../Pascal2021/dr14x.run <../Pascal2021/dr14xp.in >../Pascal2021/dr14xp.out
```

```{r code=xfun::read_utf8('../Pascal2021/dr14xp.out'), echo=TRUE, eval=FALSE}
```

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->

# Algorithm 15 - Generalized symmetric eigenproblem

We aim to solve the generalized symmetric eigenproblem

$$ A x = e B x$$

for $x$ and $e$, where symmetric matrices $A$ and $B$
and $B$ is positive definite.


## Fortran

### Listing -- Algorithm 15

```{r code=xfun::read_utf8('../fortran/alg15.f'), echo=TRUE, eval=FALSE}
```

### Example output

This example uses the Frank matrix as $B$ and a Unit matrix as matrix $A$.
Essentially we get the eigensolutions of the inverse of the Frank matrix.


```{bash run1415}
## #!/bin/bash
gfortran ../fortran/dr1415.f
mv ./a.out ../fortran/dr1415.run
../fortran/dr1415.run < ../fortran/dr1415f.in
```

<!-- ## BASIC -->

## Pascal

### Listing

```{r code=xfun::read_utf8('../Pascal2021/alg15.pas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash script15pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr15x.pas
# copy to run file
mv ../Pascal2021/dr15x ../Pascal2021/dr15x.run
../Pascal2021/dr15x.run <../Pascal2021/dr15xp.in >../Pascal2021/dr15xp.out
```


```{r code=xfun::read_utf8('../Pascal2021/dr15xp.out'), echo=TRUE, eval=FALSE}
```


<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->


# Cleanup of working files

The following script is included to remove files created during compilation or
execution of the examples.

```{bash cleanup, echo=TRUE}
## remove object and run files
cd ../fortran/
echo `pwd`
rm  *.o
rm  *.run
rm *.out
cd ../Pascal2021/
echo `pwd`
rm  *.o
rm  *.run
rm *.out
cd ../BASIC
echo `pwd`
rm *.out
cd ../Documentation
## ?? others
```


# References