---
title: "Algorithms in the Nashlib set in various programming languages -- Part 5"
author: 
   - John C Nash, retired professor, University of Ottawa
   - Peter Olsen, retired ??
date: "11/01/2021"
output: 
  pdf_document:
    keep_tex: false
    toc: true
bibliography: nashlib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## ??require(bookdown) # language engine to display text??
```

# Abstract

Algorithms 11 and 12 and 26 from the book @jncnm79 are implemented in a variety of programming
languages including Fortran, BASIC, Pascal, Python and R. 
These concern the eigensolutions of a general complex
square matrix and possible extensions.

\pagebreak
# Overview of this document

This section is repeated for each of the parts of
Nashlib documentation.

A companion document **Overview of Nashlib and its Implementations** describes the
process and computing environments for the implementation of Nashlib algorithms.
This document gives comments and/or details relating to implementations of
the algorithms themselves.

Note that some discussion of the reasoning behind certain choices in algorithms or
implementations are given in the Overview document.

\pagebreak

# Algorithm 26 -- Complex matrix eigensolutions

## Fortran 

### Listing

```{r code=xfun::read_utf8('../fortran/comeigd26.f'), echo=TRUE, eval=FALSE}
```

### Example output

?? explanation needed. Note use of Pascal input file


```{bash runf26}
gfortran ../fortran/comeigd26.f
mv ./a.out ../fortran/comeigd26.run
../fortran/comeigd26.run < ../fortran/d26f.in
```

## BASIC

### Listing

```{r code=xfun::read_utf8('../BASIC/comeiga26.bas'), echo=TRUE, eval=FALSE}
```

### Example output

```{bash a26bas}
bwbasic ../BASIC/comeiga26.bas  >../BASIC/a26.out
# echo "done"
```

```{r code=xfun::read_utf8('../BASIC/a26.out'), echo=TRUE, eval=FALSE}
```

## Pascal

### Listing

```{r code=xfun::read_utf8('../Pascal2021/dr26.pas'), echo=TRUE, eval=FALSE}
```

### Example output

We create an order 5 complex matrix where the real part is a Frank matrix and the imaginary 
part is a Moler matrix.

```{bash script2pas, eval=TRUE, echo=TRUE}
fpc ../Pascal2021/dr26.pas
# copy to run file
mv ../Pascal2021/dr26 ../Pascal2021/dr26.run
../Pascal2021/dr26.run <../Pascal2021/dr26p.in >../Pascal2021/dr26p.out
```

<!-- Note how it appears fpc output seems to have too many linefeeds (\r characters).  -->
<!-- We can use sed to strip them out. -->

```{bash strip26, echo=FALSE}
cd ../Pascal2021
sed 's/\r//' dr26p.out >tmp.out
mv tmp.out dr26p.out
```

```{r code=xfun::read_utf8('../Pascal2021/dr26p.out'), echo=TRUE, eval=FALSE}
```

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->


\pagebreak
# Algorithms 11 and 12 -- standardization and residuals for a complex eigensolution

These algorithms are probably among the least used of those included in Nashlib.
Their intent was to allow proposed eigensolutions of complex matrices to be 
standardized and tested. This seemed a potentially important task in the 1970s.
?? include COMEIG (ref to Eberlein's work, others??)

The purpose of standardization is to facilitate comparisons between eigenvectors
that are supposedly equivalent. Any (preferably) unit-length multiple of an 
eigenvector is also an eigenvector, so it is difficult to compare two proposed
solutions for the same eigenvalue. Therefore we choose a multiplier so that the
largest magnitude component of the eigenvector is set to $1 + 0 i$ where 

$$ i = \sqrt{-1}$$
## Fortran

The Algorithm 11 code:

```{r code=xfun::read_utf8('../fortran/alg11.f'), echo=TRUE, eval=FALSE}
```

The Algorithm 11 code:


```{r code=xfun::read_utf8('../fortran/alg12.f'), echo=TRUE, eval=FALSE}
```


### Example output

We illustrate by finding a single eigensolution of the Hilbert segments of
order 5 and 10. ?? Do we want to swap in the Frank matrix (the computations
are generally easier)?

```{bash run1112}
## #!/bin/bash
gfortran ../fortran/a1112.f
mv ./a.out ../fortran/a1112.run
../fortran/a1112.run <../fortran/a1112.in
```

<!-- ## BASIC -->

## Pascal

Currently we do not seem to have a separate example driver for these two codes.
However, they are used in conjunction with the example(s) for Algorithm 26 (Complex
matrix eigensolutions).

### Listing -- Algorithm 11

```{r code=xfun::read_utf8('../Pascal2021/alg11.pas'), echo=TRUE, eval=FALSE}
```

### Listing -- Algorithm 12

```{r code=xfun::read_utf8('../Pascal2021/alg12.pas'), echo=TRUE, eval=FALSE}
```

<!-- ## Python -->

<!-- ## R -->

<!-- ## Others -->







# Cleanup of working files

The following script is included to remove files created during compilation or
execution of the examples.

```{bash cleanup, echo=TRUE}
## remove object and run files
cd ../fortran/
echo `pwd`
rm  *.o
rm  *.run
rm *.out
cd ../Pascal2021/
echo `pwd`
rm  *.o
rm  *.run
rm *.out
cd ../BASIC
echo `pwd`
rm *.out
cd ../Documentation
## ?? others
```


# References